Go Implementation of Up-To Techniques for Equivalence of Weighted Languages

### 2020-09-27 paper/paper.tex
congruence closure algorithm implemented

\begin{abstract}
Weighted automata generalize non-deterministic automata by adding
a quantity expressing the weight (or probability) of the execution of each transition.
In this work we propose an implementation of two algorithms for computing language 
equivalence in finite state linear weighted automata (LWAs). The first, a
linear partition refinement algorithm, calculates the linear weighted bisimulation
for any given LWA, the second algorithm checks the equivalence 
of two vectors (states) for a given weighted automata, and is used to verify the results
of the first algorithm. We then provide and compare runtime results. 
\end{abstract}

@@

In \cite{DBLP:journals/corr/Bonchi0K17}, up-to techniques are defined for
weighted systems over arbitrary semirings, while in \cite{BONCHI201277}, up-to techniques
are studied under a more abstract coalgebraic perspective. For the sake of simplicity of 
the implementation, we focus only on weighted automata over the field of real numbers 
$\R$. Real numbers are implemented with double precision floating point numbers,  
known as the \texttt{float64} type in the Go programming language.

@@

\begin{defn}
  A \textit{weighted automaton} over a field $\K$ and an alphabet $A$ is a triple 
  $(X,o,t)$ such that $X$ is a finite set of states, $t = (t_a : X \to K)_{a \in A}$
  is a set of transition functions indexed over the symbols of the alphabet and $o : X 
  \to \K$ is the output function. The transition functions $t_a$ are represented as
  $\K^{X \times X}$ matrices. $o \in \K^X$ is represented as a row vector.
  $t_a(v)$ denotes the vector obtained by multiplying the matrix $t_a$ by the vector $v 
  \in \K^X$. $o(v)$ denotes the scalar $s \in \K^X$ obtained by multiplying the row
  vector $o$ by the column vector $v \in \K^X$.
  $A^*$ is the set of all words over $A$. $\epsilon$ is the empty word and $aw$ is the
  concatenation of a letter $a$ to the word $w \in A^*$.
  A weighted language is a function $\psi: A^* \to \K$.
  A function mapping each state vector into its 
  accepted language, $\sqmap{\cdot}: \K^X \to K^{A^*}$ is defined as follows for every weighted automaton:

@@

  Two vectors $v_1, v_2 \in \K^X$ are called language equivalent $v_1 \sim v_2 \iff \sqmap
  {v_1} = \sqmap{v_2}$. One can extend the notion of language equivalence to states rather
  than for vectors by assigning to each state $x \in X$ the corresponding unit vector 
  $e_x \in \K^X$. When given an initial state $i$ for a weighted automaton, the language 
  of the automaton can be defined as $\sqmap{i}$.
\end{defn}

@@

>>>>

\begin{abstract}
Weighted automata generalize non-deterministic automata by adding
a quantity expressing the weight (or probability) of the execution of each transition.
In this work we propose an implementation of two algorithms for computing language 
equivalence in finite state weighted automata (WAs). The first, a
linear partition refinement algorithm, calculates the largest linear weighted bisimulation
for any given LWA (Linear Weighted Automata) through an iterative method, 
the second algorithm checks the language equivalence 
of two vectors (states) for a given weighted automata by using an additional
data structure representing a congruence relationship between states.
We then compare the two algorithms results on randomly generated automata samples,
to verify their correctness and provide some runtime statistics.

\end{abstract}

@@

In \cite{DBLP:journals/corr/Bonchi0K17}, up-to techniques are defined for
weighted systems over arbitrary semirings, while in \cite{BONCHI201277}, up-to techniques
are defined for Linear Weighted Automata (LWAs), under a more abstract coalgebraic perspective.
For ease of implementation, we focus only on weighted automata over the field of real numbers 
$\R$.

@@

\begin{defn}
  A \textit{weighted automaton} over a field $\K$ and an alphabet $A$ is a triple 
  $(X,o,t)$ such that $X$ is a finite set of states, $t = (t_a : X \to K)_{a \in A}$
  is a set of transition functions indexed over the symbols of the alphabet and $o : X 
  \to \K$ is the output function. The transition functions $t_a$ are represented as
  $\K^{X \times X}$ matrices. $o \in \K^{1 \times X}$ is represented as a row vector.
  $t_a(v)$ denotes the vector obtained by multiplying the matrix $t_a$ by the column 
  vector $v  \in \K^{X \times 1}$. $o(v)$ denotes the scalar $s \in \K$ obtained by 
  multiplying the row vector $o$ by the column vector $v \in \K^{X \times 1}$.
  $A^*$ is the set of all words over $A$. $\epsilon$ is the empty word and $aw$ is the
  concatenation of a letter $a$ to the word $w \in A^*$.
  A weighted language is a function $\psi: A^* \to \K$.
  A function mapping each state vector into its 
  accepted language, $\sqmap{\cdot}: \K^X \to K^{A^*}$ is defined as follows for every weighted automaton:

@@

  Two vectors $v_1, v_2 \in \K^{X\times 1}$ are called language equivalent, denoted with $v_1 \sim v_2 \iff \sqmap
  {v_1} = \sqmap{v_2}$. One can extend the notion of language equivalence to states rather
  than for vectors by assigning to each state $x \in X$ the corresponding unit vector 
  $e_x \in \K^X$. When given an initial state $i$ for a weighted automaton, the language 
  of the automaton can be defined as $\sqmap{i}$.
\end{defn}

@@

Real numbers are implemented with double precision floating point numbers,  
known as the \texttt{float64} type in the Go programming language.
